Type Evolution in Programming Systems

### 2024-05-28 paper.tex
Replace hand-drawn diagrams

The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
we will return to repeatedly in this paper. When a type changes from $T_1$ to $T_2$,
we need to find and apply a corresponding change to other parts of the program whose shape
is defined by the changed type. The problem of \emph{type evolution} is thus finding and
applying changes that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$
that correspond to the new type.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.175]{figures/diag-data-code.png}
\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
  in data and code whose shape is defined by type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{1em}
\includegraphics[scale=0.175]{figures/diag-both.png}
\end{wrapfigure}
\frameworkboxtitle{Code-data Co-evolution}
In this case study, the evolving \emph{type} is the \texttt{State} type. When the type
evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
The former is the program logic implemented in \texttt{update} and \texttt{render},
while the latter is the current state (a live value of type \texttt{State}). The scenario
illustrates that code and data need to co-evolve as illustrated by a revised diagram on the right.
}

@@

\frameworkbox{
\begin{wrapfigure}{r}{8em}
\includegraphics[scale=0.175]{figures/diag-variants.png}
\end{wrapfigure}
\frameworkboxtitle{Primitive Type Transformations}
In this example, \emph{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

@@

\newpage
~
\newpage

>>>>

The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
we will return to repeatedly in this paper. When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$,
we need to find and apply a corresponding change to other parts of the program whose shape
is defined by the changed type. The problem of \emph{type evolution} is thus finding and
applying changes that will update $\textsf{data}_1$ and $\textsf{code}_1$ into new versions $\textsf{data}_1$ and $\textsf{code}_2$
that correspond to the new type.

@@

\begin{figure}[t]
\centering
%\includegraphics[scale=0.175]{figures/diag-data-code.png}
\begin{tikzcd}[sep = 6em]
  \textsf{code}_1
  \arrow[d, "\textsf{refactor}"{name=L, description}]
  &
  \textsf{type}_1
  \arrow[l, "\textsf{constrains}"{above}]
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}"{name=C, description}]
  &
  \textsf{data}_1
  \arrow[d, "\textsf{migrate}"{name=R, description}]
  \\
  \textsf{code}_2
  &
  \textsf{type}_2
  \arrow[l, "\textsf{constrains}"{above}]
  \arrow[r, "\textsf{constrains}"{above}]
  &
  \textsf{data}_2
  \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
\end{tikzcd}
\caption{When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$, a corresponding change needs to take place
  in data and code constrained by the type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

\frameworkbox{
% \begin{wrapfigure}{r}{14em}
% \vspace{1em}
% \includegraphics[scale=0.175]{figures/diag-both.png}
% \end{wrapfigure}
\frameworkboxtitle{Code-data Co-evolution}
In this case study, the evolving type in Figure~\ref{fig:change} is the \texttt{State} type. When the type
evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
The former is the program logic implemented in \texttt{update} and \texttt{render},
while the latter is the current state (a live value of type \texttt{State}). The scenario
illustrates that code and data need to co-evolve as illustrated by a revised diagram on the right.
}

@@

\begin{figure}[H]
\begin{tikzcd}[sep = 6em]
  \textsf{type}_1
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}_1"{name=C, description}]
  &
  \textsf{data}_1
  \arrow[d, "\textsf{migrate}_1"{name=R, description}]
  \\
  \textsf{type}_2
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
  &
  \textsf{data}_2
  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
  \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
  \\
  \textsf{type}_3
  \arrow[r, "\textsf{constrains}"{above}]
  &
  \textsf{data}_3
  \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
\end{tikzcd}
\caption{Multiversion type evolution.}
\label{fig:multiversion}
\end{figure}

\frameworkbox{
% wrapfigure doesn't seem to work with tikz
%\begin{wrapfigure}{r}{8em}
% \includegraphics[scale=0.175]{figures/diag-variants.png}
%\end{wrapfigure}
\frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} in Figure~\ref{fig:multiversion} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

@@
